---
title: "qPCR chemokines SkM"
author: "Jordi Camps"
date: "2018 M12 14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r message=FALSE}
library(readxl)
library(tidyr)
library(dplyr)
library(tibble)
library(ggplot2)
library(qpcrviia7)
library(ggpubr)
```

## Load data
```{r}
qpcr <- read_excel("data/fig1_skeletal_muscle/2016-02-09 174725_Chemokines_WT_Sgca_skMuscle_Jordi-ViiA7-export.xlsx", sheet = 3, skip = 35, col_names = TRUE)
#Change CT row in numeric
qpcr$CT <- as.numeric(qpcr$CT)
head(qpcr)
```

### Select columns to use
```{r}
qpcr <- qpcr[, c("Sample Name", "Target Name", "CT")]
```

## Annotation
### Read annotation file
```{r}
anno <- read_excel("data/fig1_skeletal_muscle/2016-02-09 174725_Chemokines_WT_Sgca_skMuscle_Jordi-annotation.xlsx", col_names = TRUE)
```

### Merge annotation file
```{r}
qpcr <- qpcr %>%
  left_join(anno, by = "Sample Name")
```

## Creating numeric CT column and change undetermined values into 40
```{r}
qpcr$CT <- as.numeric(qpcr$CT)
qpcr[is.na(qpcr$CT), "CT"] <- 40
```

## Quality control
### Samples
```{r fig.height=4, fig.width=5}
ggplot(qpcr, aes(`Sample Name`, CT)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("CT values per sample")
```

### Genes
```{r fig.height=4, fig.width=6}
ggplot(qpcr, aes(`Target Name`, CT)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("CT values per primer")
```

### Change low CT values to NA
```{r}
qpcr[qpcr$CT < 10, "CT"] <- NA
```

### Remove blank and WT4
```{r}
qpcr <- qpcr %>%
  filter(`Sample Name` != "Blanc" & `Sample Name` != "WT4")
```

### HKG
```{r fig.height=4, fig.width=4}
qpcr %>%
  filter(`Target Name` == "Rpl13a" | `Target Name` == "Psma3" | `Target Name` == "Rab35") %>%
  ggplot(aes(x = `Target Name`, y = CT, fill = genotype)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Housekeeping genes")
```

```{r}
qpcr %>%
  filter(`Target Name` == "Rpl13a" | `Target Name` == "Psma3" | `Target Name` == "Rab35") %>%
  select(`Sample Name`, `Target Name`, CT) %>%
  spread(key = `Target Name`, value = CT)
```

## Normalize to HKG
### normalize function
```{r}
normalize_hkg <- function(df, hkg) {
  temp <- NULL
  temp <- df[df$`Target Name` == hkg, ]
  temp <- temp %>%
    #filter(`Target Name` == hkg) %>%
    select(`Sample Name`, CT) %>%
    group_by(`Sample Name`) %>%
    summarize(CT_hkg = gm_mean(CT, na.rm = TRUE))
  
  temp2 <- df[df$`Target Name` != hkg, ]
  print(temp2 %>%
    group_by(`Sample Name`) %>%
    left_join(temp, by = "Sample Name") %>%
    #mutate(CT = replace(CT, is.na(CT_avg_hkg), NA)) %>%
    mutate(Delta_ct = CT_hkg - CT, rel_expr = 2^-Delta_ct) #%>%
    #arrange(Target)
    )
}
```

### Geometric mean function
```{r}
gm_mean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}
```

```{r}
qpcr <- normalize_hkg(df = qpcr, hkg = c("Rab35", "Rpl13a", "Psma3"))
```

## Plot
### Create plotting theme
```{r}
theme_jc_mult <- theme(panel.border = element_blank(), panel.grid = element_blank(), axis.line = element_line(colour = "black"),
                  axis.ticks = element_line(colour = "black"), axis.text = element_text(colour = "black"))
theme_jc_one <- theme(panel.border = element_blank(), panel.grid = element_blank(), axis.line = element_line(size = 1, colour = "black"),
                  axis.ticks = element_line(size = 1, colour = "black"), axis.text = element_text(colour = "black"))
```

```{r}
qpcr$genotype <- factor(qpcr$genotype, levels = c("Healthy", "Dystrophic"))
```

```{r}
p <- qpcr %>%
  #filter(`Target Name` != "Rpl13a" & `Target Name` != "Psma3") %>%
  ggplot(aes(x = genotype, y = Delta_ct, fill = genotype)) +
  geom_boxplot() +
  facet_wrap(~`Target Name`, scales = "free", strip.position = "left")
```

```{r fig.height=5, fig.width=10}
p + 
  scale_y_continuous("Log(Relative expression)") +
  ggthemes::scale_fill_tableau() +
  theme_bw(base_size = 14) +
  theme_jc_mult +
  theme(axis.text.x = element_blank(), strip.background = element_blank(), strip.placement = "outside", 
        strip.text = element_text(face = "italic", colour = "black"),
        axis.ticks.x = element_blank(), axis.title.x = element_blank(), axis.text.y = element_text(size = 10))
ggsave("plots/chemokines_skm/Delta_ct_per_gene.pdf", dpi = 600)
ggsave("plots/chemokines_skm/Delta_ct_per_gene.png", dpi = 600)
```

## normalize dystrophic to healthy sample
### Take average and standard deviation per Gene and Genotype
```{r}
df <- qpcr %>%
  ungroup() %>%
  group_by(`Target Name`, genotype) %>%
  summarise(replicates = n(), 
            Delta_ct_avg = mean(Delta_ct, na.rm = TRUE), 
            Delta_ct_sem = sd(Delta_ct, na.rm = TRUE)/sqrt(replicates)#,
            #Rel_expr_avg = mean(rel_expr, na.rm = TRUE),
            #Rel_expr_sem = sd(rel_expr, na.rm = TRUE)/sqrt(replicates)
            ) %>%
  mutate(Delta_ct_sem_perc = Delta_ct_sem / Delta_ct_avg * 100) %>%
  ungroup()
```

### Spread genotype in seperate columns
```{r}
df <- df %>% 
  gather(variable, value, -(`Target Name`:replicates)) %>%
  unite(temp, genotype, variable, sep = ".") %>%
  spread(temp, value)
```

### Calculate fold change and propagate sem
```{r}
df <- df %>%
  mutate(Dystrophic.Fold_change_avg = 2^(Dystrophic.Delta_ct_avg - Healthy.Delta_ct_avg),
         Dystrophic.Fold_change_sem = sqrt((Dystrophic.Delta_ct_sem_perc)^2 + (Healthy.Delta_ct_sem_perc)^2) * Dystrophic.Fold_change_avg / 100,
         Healthy.Fold_change_avg = 1,
         Healthy.Fold_change_sem = 0
         )
```

### Unite genotype again in one column
```{r}
df <- df %>%
  gather(Dystrophic.Delta_ct_avg:Healthy.Fold_change_sem, key = variable, value = value) %>%
  tidyr::separate(col = variable, into = c("Genotype", "temp1", "temp2", "temp3", "temp4")) %>%
  unite(temp1, temp2, temp3, temp4, col = "temp") %>%
  spread(key = temp, value = value)
```

### Log transform fold change and sem
```{r}
df <- df %>%
  mutate(Log_Fold_change_avg_NA = log(Fold_change_avg_NA),
         Log_Fold_change_sem_NA = Fold_change_sem_NA/(Log_Fold_change_avg_NA * log(10)))
```


### Calculate ymin and ymax
```{r}
df <- df %>%
  mutate(ymin = Log_Fold_change_avg_NA - Log_Fold_change_sem_NA, 
         ymax = Log_Fold_change_avg_NA + Log_Fold_change_sem_NA)
```

### Statistics
perform t-test dystrophic vs healthy and implement in the plot
```{r}
genes <- unique(qpcr[["Target Name"]])
t_data <- list()

for (i in 1:21){
  temp <- qpcr %>%
  filter(`Target Name` == genes[i])
t <- t.test(Delta_ct~genotype, data = temp)
t_data[[i]] <- c(p_value = t$p.value, `Target Name` = genes[i])
}
```

```{r}
stat <- as.data.frame(t(as.data.frame(t_data, stringsAsFactors = FALSE)), stringsAsFactors = FALSE)
stat$p_value <- as.numeric(stat$p_value)
stat
```

```{r}
df <- df %>%
  left_join(stat, by = "Target Name")
```

### Create boolean for p-value and prepare plotting
```{r}
p <- df %>%
  mutate(sign = p_value <= 0.05) %>%
  ungroup() %>%
  filter(`Target Name` != "Rpl13a" & `Target Name` != "Rab35") %>%
  filter(Genotype == "Dystrophic") %>%
  arrange(desc(Fold_change_avg_NA)) %>%
  mutate(order = row_number())
```

### Plot
```{r fig.height=5, fig.width=2.5}
p %>%
  ggplot(aes(x = order, y = rev(Log_Fold_change_avg_NA), fill = rev(sign))) +
    geom_col(position = position_dodge(), width = 0.8) +
    geom_errorbar(aes(ymin = rev(ymin), ymax = rev(ymax)), position = position_dodge(0.9), width = 0.25) +
    geom_hline(yintercept = log(1), color = "red", size = 1) +
    scale_fill_manual(values = c("#bababa", "#1a1a1a"), labels = c("p > 0.05", "p < 0.05")) +
    scale_y_continuous("Log(fold change) Dystrophic\nvs Healthy skeletal muscle", expand = c(0, 0.1)) +
    scale_x_continuous(breaks = p$order, labels = rev(p$`Target Name`), expand = c(0, 0)) +
    coord_flip() +
    theme_bw(base_size = 12) +
    theme_jc_one +
    theme(axis.title.y = element_blank(), legend.position = "top", legend.title = element_blank(), 
          axis.text.y = element_text(size = 12, face = "italic"), axis.ticks.y = element_blank(), axis.title.x = element_text(hjust = 1))
ggsave("plots/chemokines_skm/fold_change_Dys_vs_wt.pdf", dpi = 600)
```

## Plot only significant genes
### Equal breaks function
```{r}
equal_breaks <- function(n = 3, s = 0.05, ...){
  function(x){
    # rescaling
    d <- s * diff(range(x)) / (1+2*s)
    round(seq(min(x)+d, max(x)-d, length=n), 1)
  }
}
```

### Select siginificant genes and order df according to highest fold change
```{r fig.height=4, fig.width=6}
temp <- p %>%
  filter(sign == TRUE) %>%
  select(`Target Name`, sign, Fold_change_avg_NA) %>%
  inner_join(qpcr) %>%
  ungroup() %>%
  arrange(desc(Fold_change_avg_NA))
temp$`Target Name` <- factor(temp$`Target Name`, levels = unique(temp$`Target Name`))
```

```{r fig.height=5, fig.width=5}
temp %>%
  ggplot(aes(x = genotype, y = Delta_ct, fill = genotype)) +
  geom_boxplot() +
  facet_wrap(~`Target Name`, scales = "free", strip.position = "left") +
  scale_y_continuous("Delta CT", breaks = equal_breaks(n = 4, s = 0.05), expand = c(0.05, 0)) +
  ggthemes::scale_fill_tableau() +
  theme_bw(base_size = 14) +
  theme_jc_mult +
  theme(axis.text.x = element_blank(), strip.background = element_blank(), strip.placement = "outside", 
        strip.text = element_text(face = "italic", colour = "black"), legend.position = "top", legend.title = element_blank(),
        axis.ticks.x = element_blank(), axis.title.x = element_blank(), axis.text.y = element_text(size = 9))
ggsave("plots/chemokines_skm/Delta_ct_per_significant_genes.pdf", dpi = 600)
ggsave("plots/chemokines_skm/Delta_ct_per_significant_genes.png", dpi = 600)
```

## Plot total chemokines per genotype
```{r fig.height=1.25, fig.width=3}
qpcr %>%
  as_tibble() %>%
  group_by(genotype, `Target Name`) %>%
  summarize(avg_Delta_Ct = mean(Delta_ct, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(genotype, levels = c("Dystrophic", "Healthy")) , y = avg_Delta_Ct, fill = genotype)) +
  geom_boxplot(notch = TRUE, size = .6, color = "black") +
  scale_y_continuous("Delta Ct") +
  #scale_x_reverse() +
  scale_fill_manual(values = c("#67a9cf", "#ef8a62")) +
  coord_flip() +
  stat_compare_means(method = "t.test", label = "p.signif", label.x = 1.5) +
  theme_bw(base_size = 12) +
  theme(panel.border = element_blank(), panel.grid = element_blank(), axis.line = element_line(colour = "black"), 
        axis.ticks = element_line(colour = "black"), axis.text = element_text(colour = "black"), axis.title.y = element_blank(), legend.position = "none",
        axis.text.y = element_text(angle = 0, hjust = 1))
ggsave("plots/chemokines_skm/total_expression_genotype.pdf", dpi = 600)
ggsave("plots/chemokines_skm/total_expression_genotype.png", dpi = 600)
```


### Select specific genes and plot
#### CC chemokines
```{r}
genes <- c("Ccl3", "Ccl4", "Ccl12", "Ccl7", "Ccl8", "Ccl17", "Ccl22", "Ccl24", "Ccl5", "Ccl6", "Ccl9", "Ccl28", "Ccl25")
cc <- qpcr[qpcr$`Target Name` %in% genes, ]
order <- tbl_df(cc) %>%
  filter(genotype == "Healthy") %>%
  group_by(`Target Name`, genotype) %>%
  #top_n(n = 5, wt = Delta_ct) %>%
  summarise(avg_delta_ct = median(Delta_ct, na.rm = TRUE)) %>%
  arrange(avg_delta_ct) %>%
  ungroup() %>%
  mutate(order = row_number()) %>%
  select(`Target Name`, order)
```


```{r fig.height=2.75, fig.width=4.5}
cc <- cc %>%
  left_join(order)
cc %>%
  ggplot(aes(x = factor(order), y = Delta_ct, fill = genotype)) +
  geom_boxplot(col = "black") +
  #facet_wrap(~`Target Name`, scales = "free", strip.position = "top") +
  scale_y_continuous("Delta CT") +
  scale_x_discrete(breaks = cc$order, labels = cc$`Target Name`) +
  stat_compare_means(label = "p.signif", method = "t.test", hide.ns = TRUE) +
  #ggthemes::scale_fill_tableau() +
  scale_fill_manual(values = c("#67a9cf", "#ef8a62")) +
  theme_bw(base_size = 12) +
  theme_jc_mult +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"), strip.background = element_blank(), strip.placement = "outside", 
        legend.position = "top", legend.title = element_blank(), plot.margin = unit(c(0, 0, 0, 0), "lines"),
        axis.ticks.x = element_line(), axis.title.x = element_blank(), axis.text.y = element_text())
ggsave("plots/chemokines_skm/cc_chemokines.pdf", dpi = 600)
ggsave("plots/chemokines_skm/cc_chemokines.png", dpi = 600)
```

#### CXC and CX3C chemokines
```{r}
genes <- c("Cx3cl1", "Sdf1")
cxc <- temp[temp$`Target Name` %in% genes, ]
order <- tbl_df(cxc) %>%
  filter(genotype == "Healthy") %>%
  group_by(`Target Name`, genotype) %>%
  #top_n(n = 5, wt = Delta_ct) %>%
  summarise(avg_delta_ct = median(Delta_ct, na.rm = TRUE)) %>%
  arrange(avg_delta_ct) %>%
  ungroup() %>%
  mutate(order = row_number()) %>%
  select(`Target Name`, order)
```

```{r fig.height=2.75, fig.width=1.75}
cxc <- cxc %>%
  left_join(order)
cxc %>%
  ggplot(aes(x = factor(order), y = Delta_ct, fill = genotype)) +
  geom_boxplot(col = "black") +
  #facet_wrap(~`Target Name`, scales = "free", strip.position = "top") +
  scale_y_continuous("Delta CT", expand = c(0.05, 0)) +
  scale_x_discrete(breaks = cxc$order, labels = cxc$`Target Name`) +
  stat_compare_means(label = "p.signif", method = "t.test", hide.ns = TRUE) +
  #ggthemes::scale_fill_tableau() +
  scale_fill_manual(values = c("#67a9cf", "#ef8a62")) +
  theme_bw(base_size = 12) +
  theme_jc_mult +
  theme(axis.text.x = element_text(face = "italic", angle = 45, hjust = 1), strip.background = element_blank(), strip.placement = "outside", 
        strip.text = element_text(face = "italic", colour = "black"), legend.position = "top", legend.title = element_blank(),
        axis.ticks.x = element_line(), axis.title.x = element_blank(), axis.text.y = element_text(), plot.margin = unit(c(0, 0, 0, 0), "lines"))
ggsave("plots/chemokines_skm/cxc_chemokines.pdf", dpi = 600)
ggsave("plots/chemokines_skm/cxc_chemokines.png", dpi = 600)
```

#### Growth factors
```{r}
genes <- c("Pdgfa", "Pdgfb")
gf <- temp[temp$`Target Name` %in% genes, ]
order <- tbl_df(gf) %>%
  filter(genotype == "Healthy") %>%
  group_by(`Target Name`, genotype) %>%
  #top_n(n = 5, wt = Delta_ct) %>%
  summarise(avg_delta_ct = median(Delta_ct, na.rm = TRUE)) %>%
  arrange(avg_delta_ct) %>%
  ungroup() %>%
  mutate(order = row_number()) %>%
  select(`Target Name`, order)
```

```{r fig.height=2.75, fig.width=1.75}
gf <- gf %>%
  left_join(order)
gf %>%
  ggplot(aes(x = factor(order), y = Delta_ct, fill = genotype)) +
  geom_boxplot(col = "black") +
  #facet_wrap(~`Target Name`, scales = "free", strip.position = "top") +
  scale_y_continuous("Delta CT", expand = c(0.05, 0)) +
  scale_x_discrete(breaks = gf$order, labels = gf$`Target Name`) +
  stat_compare_means(label = "p.signif", method = "t.test", hide.ns = TRUE) +
  #ggthemes::scale_fill_tableau() +
  scale_fill_manual(values = c("#67a9cf", "#ef8a62")) +
  theme_bw(base_size = 12) +
  theme_jc_mult +
  theme(axis.text.x = element_text(face = "italic", angle = 45, hjust = 1), strip.background = element_blank(), strip.placement = "outside", 
        strip.text = element_text(face = "italic", colour = "black"), legend.position = "top", legend.title = element_blank(),
        axis.ticks.x = element_line(), axis.title.x = element_blank(), axis.text.y = element_text(), plot.margin = unit(c(0, 0, 0, 0), "lines"))
ggsave("plots/chemokines_skm/pdgf_chemokines.pdf", dpi = 600)
ggsave("plots/chemokines_skm/pdgf_chemokines.png", dpi = 600)
```

## Combine fold change and delta CT
```{r fig.height=3, fig.width=3}
p %>%
  ggplot(aes(x = Delta_ct_avg_NA, y = log(Fold_change_avg_NA), label = `Target Name`)) +
  #geom_point() +
  geom_text() +
  theme_bw(base_size = 12)
```

```{r fig.height=4.5, fig.width=3}
temp <- p %>%
  arrange(Fold_change_avg_NA) %>%
  mutate(order = row_number())

temp %>%
  ggplot(aes(x = order, y = log(Fold_change_avg_NA))) +
  geom_linerange(aes(x = order, ymin = 0, ymax = log(Fold_change_avg_NA)), linetype = 3, color = "slategrey") +
  geom_point(aes(col = sign, size = Delta_ct_avg_NA)) +
  scale_x_continuous(breaks = temp$order, labels = temp$`Target Name`, expand = c(0, 0.5)) +
  scale_y_continuous("Log(Fold change)", limits = c(0,5), expand = c(0, 0)) +
  scale_color_manual(values = c("slategrey", "#ef8a62"), labels = c("P > 0.05", "P < 0.05")) +
  scale_size_continuous("Delta Ct") +
  coord_flip() +
  guides(colour = guide_legend(title = "", order = 1), size = guide_legend(order = 2)) +
  theme_bw(base_size = 12) +
  theme(panel.border = element_blank(), panel.grid = element_blank(), axis.line = element_line(colour = "black"), 
        axis.ticks = element_line(colour = "black"), axis.text = element_text(colour = "black"), axis.title.y = element_blank(),
        axis.text.y = element_text(face = "italic"), legend.justification = c(1, 0.1), legend.position = c(1, 0.1), 
        plot.margin = unit(c(0, 0, 0, 0), "lines"))
ggsave("plots/chemokines_skm/bubble_chart.pdf", dpi = 600)
ggsave("plots/chemokines_skm/bubble_chart.png", dpi = 600)
```

